ForOpenCL: Transformations Exploiting Array Syntax in Fortran for Accelerator Programming

### 2011-06-30 papers/GPUScA-2011/programming-model.tex
reference, and some compression

A number of previous efforts have exploited data-parallel programming at the
language level to utilize novel architectures.
%%, particularly in previous decades during the reign of vector and massively parallel computers in high performance computing (HPC). 
The origin of the array syntax adopted by Fortran in the 1990
standard can be found in the APL language~\cite{iverson79apl}.  These additions
to Fortran allowed parallelism to be expressed with whole-array operations at
the expression level, instead of via parallelism within explicit DO-loops, as
implemented in earlier variants of the language (e.g.,
IVTRAN for the Illiac IV).

The High Performance Fortran (HPF) extension of Fortran was proposed to add
features to the language that would enhance the ability of compilers to emit
fast parallel code for distributed and shared memory parallel
computers\cite{koelbel94hpf}.  One of the notable additions to the language in
HPF was syntax to specify the distribution of data structures amongst a set of
parallel processors.  HPF also introduced an alternative looping construct to
the traditional DO-loop called {\tt FORALL} that was better suited for parallel
compilation.  An additional keyword, {\tt INDEPENDENT}, was added to allow the
programmer to indicate when the loop contained no loop-order dependencies that
allowed for parallel execution.  These constructs are similar to {\tt DO
  CONCURRENT}, an addition to Fortran in 2008.  Interestingly, the parallelism
features introduced in HPF did not exploit the new array features introduced in
1990 in any significant way, relying instead on explicit loop-based parallelism.
This restriction allowed the language to support parallel programming that
wasn't easily mapped onto a pure data-parallel model.

@@

Unfortunately, this had
two side effects that have limited broad acceptance of the whole-array
programming model in Fortran.  First, programmers must translate their
algorithms into a set of global operations.  Finite difference stencils and
similar computations are traditionally defined in terms of offsets from some
central index.  Shifting, while conceptually analogous, can be awkward to think
about for high dimensional stencils with many points.  Second, the semantics of
these operations are such that all elements of an array operation are updated as
if they were updated simultaneously.  In a program where the programmer
explicitly manages arrays and loops, double buffering techniques and user
managed temporaries are used to maintain these semantics.  When the compiler is
responsible for managing this intermediate storage, it has historically proven
that they are inefficient and generate code that requires far more temporary
storage than really necessary.  This is not a flaw of the language constructs,
but a sign of the lack of sophistication of the compilers with respect to their
internal analysis to determine how to optimally generate this intermediate
storage.

>>>>

A number of previous efforts have exploited data-parallel programming
at the language level to utilize novel architectures.  The origin of
the array syntax adopted by Fortran in the 1990 standard can be found
in the APL language~\cite{iverson79apl}.  These additions to Fortran
allowed parallelism to be expressed with whole-array operations at the
expression level, instead of via parallelism within explicit DO-loops,
as implemented in earlier variants of the language (e.g., IVTRAN for the
Illiac IV).

The High Performance Fortran (HPF) extension of Fortran was proposed
to add features to the language that would enhance the ability of
compilers to emit fast parallel code for distributed and shared memory
parallel computers\cite{koelbel94hpf}.  One of the notable additions
to the language in HPF was syntax to specify the distribution of data
structures amongst a set of parallel processors.  HPF also introduced
an alternative looping construct to the traditional DO-loop called
{\tt FORALL} that was better suited for parallel compilation.  An
additional keyword, {\tt INDEPENDENT}, was added to allow the
programmer to indicate when the loop contained no loop-order
dependencies that allowed for parallel execution.  These constructs
are similar to {\tt DO CONCURRENT}, an addition to Fortran in 2008.

Interestingly, the parallelism features introduced in HPF did not
exploit the new array features introduced in 1990 in any significant
way, relying instead on explicit loop-based parallelism.  This
restriction allowed the language to support parallel programming that
wasn't easily mapped onto a pure data-parallel model.  The {\tt SHADOW}
directive introduced in HPF-2, and the {\tt HALO} in HPF+~\cite{benkner99hpf}
bear some similarity to the halo region concept that we discuss in this 
paper.

@@

Unfortunately, this had two side effects that have limited broad
acceptance of the whole-array programming model in Fortran.  First,
programmers must translate their algorithms into a set of global
operations.  Finite difference stencils and similar computations are
traditionally defined in terms of offsets from some central index.
Shifting, while conceptually analogous, can be awkward to think about
for high dimensional stencils with many points.  Second, the semantics
of these operations are such that all elements of an array operation
are updated as if they were updated simultaneously.  In a program
where the programmer explicitly manages arrays and loops, double
buffering techniques and user managed temporaries are used to maintain
these semantics.  Limited attention to optimizing memory usage due to
this intermediate storage by compilers has led to these constructs seeing
little adoption by programmers.

%When the compiler is responsible for managing this
%intermediate storage, it has historically proven that they are
%inefficient and generate code that requires far more temporary storage
%than really necessary.  
%This is not a flaw of the language constructs,
%but a sign of the lack of sophistication of the compilers with respect
%to their internal analysis to determine how to optimally generate this
%intermediate storage.
